Fractional Types

### 2019-10-31 fracGC/p.tex
more on intro

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
combination of the Toffoli and Hadamard gates for quantum computing
should not distract from efficiency and safety concerns. The theorems
proving universality (i) assume that an unlimited amount of temporary
storage may be
used~\cite{Toffoli:1980,DBLP:conf/innovations/AaronsonGS17}, and (ii)
that this temporary storage is returned to its initial state before
de-allocation. The latter condition is necessary in order maintain
symmetry between input and output in reversible circuits and to avoid
decoherence problems in quantum computing, Unfortunately, in current
proposals for reversible and quantum programming languages
(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this condition
is at the programmer's discretion.

\paragraph*{Negative Entropy.}  
According to the conventional theory of
communication~\cite{Shannon1948}, a type with $N$ values is viewed as
an abstract system that has $N$ distinguishable states where the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The logarithmic map
implies that information contained in a composite state is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
the information contained in each subsystem. A \emph{fractional type}
$\frac{1}{A}$ introduces negative entropy. For example, a type with
cardinality $\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is
natural to interpret this negative entropy just like we interpret
``negative money,'' as a resource (space or bandwidth) to be repaid
(reclaimed) by some other part of the system.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivating Example}

>>>>

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
combination of the Toffoli and Hadamard gates for quantum computing
should not distract from efficiency and safety concerns. The theorems
proving universality (i) assume that temporary storage (often called
\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
this temporary storage is returned to its initial state before
de-allocation. Indeed if no temporary storage is allowed, the Toffoli
gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
we demonstrate using space-time tradeoffs in Sec.~\ref{sec:examples},
the more temporary storage is allowed, the more efficient certain
computations could become. The condition requiring that the temporary
storage is only de-allocated when returned to its initial condition is
a safety condition. Violating this condition destroys the symmetry
between input and output making the circuits not reversible and, in
the quantum model, causes irreversible decoherence problems. As
reviewed in Sec.~\ref{sec:examples}, current proposals for reversible
and quantum programming languages
(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this safety
condition is at the programmer's discretion whereas.

\paragraph*{Negative Entropy.}  
According to the conventional theory of
communication~\cite{Shannon1948}, a type with $N$ values is viewed as
an abstract system that has $N$ distinguishable states where the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The logarithmic map
implies that information contained in a composite state is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
the information contained in each subsystem. A \emph{fractional type}
$\frac{1}{A}$ introduces negative entropy. For example, a type with
cardinality $\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is
natural to interpret this negative entropy just like we interpret
``negative money,'' as a resource (space or bandwidth) to be repaid
(reclaimed) by some other part of the system. Indeed, we will
introduce such fractional types and use them to represent ``garbage
collection processes'' that reclaim temporary storage.

\paragraph*{Contributions.}

\paragraph*{Outline.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivating Examples}
\label{sec:examples}

Toffoli4 using 2 Toffoli3: core of proof of universality; simple
enough

In-place matrix transpose: ease of programming, efficiency

Example(s) from Quipper etc. with focus on safety condition
